\documentclass[12pt]{amsart}

\newif\ifdraft
\drafttrue

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}

\usepackage{lmodern}
\usepackage{microtype}
\usepackage{fullpage}

\usepackage{newunicodechar}
\newunicodechar{ψ}{\psi}

\input{Qcircuit}
\let\bra\relax
\let\ket\relax
\usepackage{braket}

\usepackage{mathtools}

\usepackage[svgnames]{xcolor}

\usepackage{tikz}
\usetikzlibrary{calc}

\newcommand{\asz}[1]{\ifdraft\textcolor{violet}{[ASZ: #1]}\fi}
\newcommand{\jp}[1]{\ifdraft\textcolor{orange}{[JP: #1]}\fi}

\newcommand*{\bit}{\mathbb{B}}
\newcommand*{\qubit}{\mathbb{Q}}
\newcommand*{\tensor}{\otimes}

\newcommand*{\term}[1]{\emph{#1}}

\newcommand*{\latin}[1]{\textit{#1}}
\newcommand*{\IE}{\latin{i.e.}}

\usepackage[section]{placeins} % Floats should stay in their section

\usepackage[ colorlinks
           , pagebackref
           , linktoc = all
           , pdfusetitle ]{hyperref}
\usepackage[all]{hypcap}
\renewcommand*{\backrefalt}[4]{%
  \ifcase #1 %
    No citations.%
  \or
    Cited on page #2.%
  \else
    Cited on pages #2.%
  \fi
}
\hypersetup{ linkcolor = DarkBlue
           , citecolor = DarkGreen
           , urlcolor  = DarkGoldenrod
           , filecolor = Orange
           , menucolor = DarkRed
           , runcolor  = OrangeRed }

% Cleveref (must be loaded late---after hyperref, no less!)
\usepackage[noabbrev,capitalize,nameinlink]{cleveref}
\newcommand{\creflastconjunction}{, and~}

\begin{document}

\section{Quantum computation}\label{sec:quantum-computation}

When we change from classical\footnote{By \term{classical} here, we do not mean
``affirming the law of the excluded middle'', as in classical logic
 s.\ intuitionistic logic; we mean ``following the laws of traditional Newtonian
physics'', as in classical physics vs.\ quantum mechanics.  This naming
collision is unfortunate; however, in this paper, we do not discuss
classical-logic--inspired computation, and so the term will always refer to the
physical concept.\asz{This probably belongs earlier.}} to quantum computation,
our primitive data type changes from the classical bit to the quantum
\term{qubit} (short for \emph{qu}antum \emph{bit}).  At first glance, qubits
behave much like bits: just as measuring a bit produces one of the two values
$0$ or $1$, measuring a qubit produces one of the two values $\ket{0}$ or
$\ket{1}$.  However, qubits behave very differently when \emph{not} being
measured.  This is because qubits exhibit two properties that classical bits do
not:\asz{\term{superposition} and \term{entanglement}. -- does this belong
here?}
\begin{enumerate}
  \item \term{Superposition}: A qubit, prior to measurement, can be a
    \emph{combination} of $\ket{0}$ and $\ket{1}$.

  \item \term{Entanglement}: Two (or more) qubits can be related in such a way
    that measuring the value of the first provides information about the value
    of the second.\asz{This explanation doesn't make clear why entanglement is
    magic, and I don't really know how to explain this without the math.  I'll
    come back to this.}
\end{enumerate}

To explore the computational properties these give rise to, we will explore four
different algorithms that require more of these features:
\begin{enumerate}
  \item Bitwise negation (\cref{sec:bitwise-negation-hl}), which is classical;
  \item Simulating a fair coin (\cref{sec:fair-coin-hl}), which requires
    superposition;
  \item Agreeing on the result of a fair coin (\cref{sec:agree-coin-hl}), which
    requires entanglement; and
  \item Quantum teleportation (\cref{sec:teleportation-hl}), which makes
    nontrivial use of both superposition and entanglement.
\end{enumerate}

\subsection{Bitwise negation}\label{sec:bitwise-negation-hl}
Given a qubit (or, equally well, a classical bit), we can send it through a not
gate, $X$, which changes $\ket{0}$ into $\ket{1}$ and $\ket{1}$ into $\ket{0}$;
the quantum circuit diagram\asz{We need to explain how to read these.} for this
is in \cref{fig:bitwise-negation}, demonstrating the bitwise negation of
$\ket{0}$.

\begin{figure}
  \centerline{\Qcircuit{
    \lstick{\ket{0}} & \gate{X} & \qw & \lstick{\ket{1}}
  }}
  \caption{Bitwise negation (of $\ket{0}$)}\label{fig:bitwise-negation}
\end{figure}

\subsection{Simulating a fair coin}\label{sec:fair-coin-hl}
As we said above, a qubit may be in a \term{superposition} of $\ket{0}$ and
$\ket{1}$; such states are called \term{mixed states}.  The simplest
superposition is one which is ``$50\%$ $\ket{0}$ and $50\%$ $\ket{1}$''; to
produce this, we use the quantum \term{Hadamard gate} $H$, which takes both
$\ket{0}$ and $\ket{1}$ to such mixed states.\footnote{It takes them to two
\emph{different} such states -- see \asz{some later section}.}  Then, when this
mixed state is measured, the measurement is \emph{probabilistic}: it has a 50\%
chance of being $\ket{0}$ and a 50\% chance of being $\ket{1}$.
\jp{More should be said about measurement.}


\subsection{Agreeing on the result of a fair coin}\label{sec:agree-coin-hl}
\asz{Maybe we only need this (and the preceding two) at the concrete level.}

\subsection{Quantum teleportation}\label{sec:teleportation-hl}
Quantum teleportation allows us to do something very special: given some shared
state between two parties, sending only two classical bits will allow us to
transmit an entire qubit (\IE, two complex numbers).  Suppose Alice wishes to
send a single qubit to Bob.  Then the algorithm proceeds as follows.

\begin{enumerate}
  \item \emph{Preparation.} Alice and Bob meet, and entangle two qubits, $a$ and
    $b$, so that they are guaranteed to be equal but which qubit they will be
    measured as is arbitrary.  Then they leave; Alice takes qubit $a$ with her,
    and Bob takes qubit $b$ with him.

  \item \emph{Composition.}  Sometime later, Alice places her message qubit $m$
    into some arbitrary state $\ket{\psi}$.\asz{I \emph{think} kets aren't only
    used for pure states.}

  \item \emph{Measurement.}  Alice now wishes to send $\psi$ to Bob.  In order
    to do this, Alice entangles $m$ and $a$ via a rotation\asz{check this
    terminology} of the two: all pure states ($\ket{00}$, $\ket{01}$,
    $\ket{10}$, and $\ket{11}$) are transformed into superpositions of two
    different pure states.  Since $a$ was already entangled with Bob's qubit
    $b$, this means that $m$ is entangled with $b$ as well.\asz{Should we split
    this next off into a separate step?}  Alice then measures $a$ and $b$,
    getting the two classical bits $m_M$ and $a_M$.

  \item \emph{Classical communication.}  Alice sends Bob the classical message
    $(m_M,a_M)$, via any arbitrary classical channel.

  \item \asz{I decided this isn't clear without the details.  Maybe I'll try
    explaining this to random victims -- er, friends -- and see how clear I can
    make it.} 
\end{enumerate}

\section{Quantum teleportation}\label{sec:teleportation}

There are at least three ways to think about the quantum teleportation
algorithm.  The first is to say that it's a method for transfering an arbitrary
quantum state from one person two another; this focuses on the ``teleportation''
part.  The second is to say that it's a way to send two complex numbers (a
quantum state) to another person by sending them two bits, as long as you share
some pre-arranged material (an entangled state); this focuses on the way that
quantum computation allows unusually efficient computation.  The third way is to
say that quantum teleportation is the construction of ``a `single-use
isomorphism' between the (otherwise isomorphic) types [of a qubit] and [of two
bits]''  (Selinger and Valiron, 2009\asz{reify this citation}); that is, it
allows us to turn a qubit into two bits, and turn those two bits back into a
qubit – but only \emph{once}.

We now walk through a standard approach to the quantum teleportation algorithm,
as seen in Abramsky and Coecke (2008) and Selinger and Valiron (2009)\asz{reify
these citations}.  The algorithm's quantum circuit presentation is given in
\cref{fig:quantum-teleportation-circuit}

\begin{figure}
\def\Qm{0}
\def\Qa{-2}
\def\Qb{-4}
\begin{tikzpicture}
  \def\labeledket#1#2{\mathllap{#1 \qquad} \ket{#2}}
  \tikzset{wire label/.style={minimum width=1.5em}}
  
  \tikzset{gate/.style={rectangle, minimum width=1em, draw=black, fill=white}}
  \def\cgateradius{4pt}
  
  \def\hadamard at#1;{
    \node[gate] at #1 {$H$} ;
  }

  \def\control at#1;{
    \filldraw[black] #1 circle (\cgateradius) ;
  }
  
  \def\cnot at#1from#2to#3;{
    \control at  (#1,#2) ;
    \draw[black] (#1,#3) circle (\cgateradius) ;
    \draw[black] (#1,#2) -- ($(#1,#3)+(0,-\cgateradius)$) ;
  }
  
  \def\wire at#1for#2from#3to#4;{
    \draw (0,#1)
          node[wire label, left]  {$\labeledket{#2}{#3}$}
          --
          (14,#1)
          node[wire label, right] {$\ket{#4}$} ;
  }
  
  \tikzset{cwire/.style={double distance=2pt}}

  \def\send cbit at#1from#2to#3;{
    \draw[cwire] (#1,#2) -- (#1,#3) ;
    \control at (#1,#2) ;
  }
  
  \wire at \Qm for m from ψ to x ;
  \wire at \Qa for a from 0 to y ;
  \wire at \Qb for b from 0 to ψ ;
  
  \hadamard at (1.5,\Qa) ;
  \cnot at 3 from \Qa to \Qb ;
  
  \cnot at 5 from \Qm to \Qa ;
  \hadamard at (6.5,\Qm) ;

  \pgfmathsetmacro\Qma{((\Qm)+(\Qa))/2}
  \pgfmathsetmacro\meassize{ 2cm               % gate width
                           + 1.5em             % extra for plain gate size
                           - 1.4pt - 0.35 pt } % line width adjustment
  \pgfmathsetmacro\meassizehalf{\meassize/2}
  \pgfmathsetmacro\measwidthflat{2.5cm - \meassizehalf pt}

  \draw[cwire] (9,\Qm) -- (14,\Qm) ;
  \draw[cwire] (9,\Qa) -- (14,\Qa) ;
  
  \path[fill=white,draw=black]
    ($(8,\Qma) + (0,\meassizehalf pt)$)
    -- ++(\measwidthflat pt,0)
    arc (90:-90:\meassizehalf pt)
    -- ++(-\measwidthflat pt,0)
    -- cycle ;

  \coordinate (meas meter flat mid)
    at ($(8,\Qma) + (\meassizehalf pt, -0.4em) + (-0.25em,0)$) ;
  \draw (meas meter flat mid) ++(-2em,0) .. controls ++(2em,1em) .. ++(4em,0)
        (meas meter flat mid) ++(-0.375em,0) -- ++(0.75em,1.5em) ;

  \send cbit at 12.25 from \Qm to \Qb ;
  \send cbit at 12.75 from \Qa to \Qb ;
  \node[gate] at (12.5,\Qb) {$F_{x,y}$} ;
\end{tikzpicture}
\caption{The quantum circuit diagram for quantum teleportation.}%
\label{fig:quantum-teleportation-circuit}
\end{figure}

\section{Quantum teleportation, with matrices}\label{sec:teleportation-matrices}

\asz{Trying again}

Quantum teleportation allows us to move a $1$-qubit state from Alice to Bob
simply by sending two classical bits, as long as Alice and Bob share an
entangled pair of qubits.  The algorithm is as follows.

\begin{enumerate}
  \item \emph{Preparation.}  Alice and Bob meet, and entangle two qubits, $a$
    and $b$, which were initially each in the state $\ket{0}$.  These qubits
    are entangled into the state \[ \frac{\ket{00} + \ket{11}}{\sqrt{2}} \] by
    the operation
    \begin{align*}
      P
      &= C(H \tensor I) \\
      &= C
         \left[ \frac{1}{\sqrt{2}}
                \begin{pmatrix} 1 & 1 \\ -1 & -1 \end{pmatrix}
                \tensor
                \begin{pmatrix} 1 & 0 \\  0 &  1 \end{pmatrix} \right] \\
      &= C
         \frac{1}{\sqrt{2}}
         \begin{pmatrix}
           1 & 0 &  1 &  0 \\
           0 & 1 &  0 &  1 \\
           1 & 0 & -1 &  0 \\
           0 & 1 &  0 & -1
         \end{pmatrix} \\
      &= \begin{pmatrix}
           1 & 0 & 0 & 0 \\
           0 & 1 & 0 & 0 \\
           0 & 0 & 0 & 1 \\
           0 & 0 & 1 & 0
         \end{pmatrix}
         \frac{1}{\sqrt{2}}
         \begin{pmatrix}
           1 & 0 &  1 &  0 \\
           0 & 1 &  0 &  1 \\
           1 & 0 & -1 &  0 \\
           0 & 1 &  0 & -1
         \end{pmatrix} \\
      &= \frac{1}{\sqrt{2}}
         \begin{pmatrix}
           1 & 0 &  1 &  0 \\
           0 & 1 &  0 &  1 \\
           0 & 1 &  0 & -1 \\
           1 & 0 & -1 &  0
         \end{pmatrix}.
    \end{align*}
    Thus, we have \[ P\ket{00} = P\begin{pmatrix} 1 \\ 0 \\ 0 \\ 0 \end{pmatrix}
    = \frac{1}{\sqrt{2}} \begin{pmatrix} 1 \\ 0 \\ 0 \\ 1 \end{pmatrix} =
    \frac{\ket{00} + \ket{11}}{\sqrt{2}}, \] as we wanted.

  \item \emph{Composition.}  At some later time, Alice places her message qubit
    $m$ into some arbitrary state $\ket{ψ}$.

  \item \emph{Measurement.}  Alice now wishes to send the state $\ket{ψ}$ to
    Bob.  To do this, Alice measures qubits $m$ and $a$ in a \emph{different}
    basis, which involves entangling $m$ and $a$ before measuring them.  This
    means that $m$ is entangled with $b$ as well, which is key.  Entanglement of
    $m$ and $a$ is the reverse of the entanglement of $a$ and $b$:
    \begin{align*}
      M
      &= (H \tensor I)C \\
      &= \left[ \frac{1}{\sqrt{2}}
                \begin{pmatrix} 1 & 1 \\ -1 & -1 \end{pmatrix}
                \tensor
                \begin{pmatrix} 1 & 0 \\  0 &  1 \end{pmatrix} \right]
         C \\
      &= \frac{1}{\sqrt{2}}
         \begin{pmatrix}
           1 & 0 &  1 &  0 \\
           0 & 1 &  0 &  1 \\
           1 & 0 & -1 &  0 \\
           0 & 1 &  0 & -1
         \end{pmatrix}
         C\\
      &= \frac{1}{\sqrt{2}}
         \begin{pmatrix}
           1 & 0 &  1 &  0 \\
           0 & 1 &  0 &  1 \\
           1 & 0 & -1 &  0 \\
           0 & 1 &  0 & -1
         \end{pmatrix}
         \begin{pmatrix}
           1 & 0 & 0 & 0 \\
           0 & 1 & 0 & 0 \\
           0 & 0 & 0 & 1 \\
           0 & 0 & 1 & 0
         \end{pmatrix} \\
      &= \frac{1}{\sqrt{2}}
         \begin{pmatrix}
           1 & 0 &  0 &  1 \\
           0 & 1 &  1 &  0 \\
           1 & 0 &  0 & -1 \\
           0 & 1 & -1 &  0
         \end{pmatrix}
    \end{align*}
    Since $H^2 = I$, $I^2 = I$, and $C^2 = I$, we know that $M = P^{-1}$ (which
    we may also see by noting that $M = P^\dag = P^{\mathsf{T}}$, and $M$ and
    $P$ are unitary).  Thus, we know that $M$ measures in the following basis:
    \begin{alignat*}{4}
      M\frac{\ket{00} + \ket{11}}{\sqrt{2}}
        &= M\frac{1}{\sqrt{2}}\begin{pmatrix} 1 \\ 0 \\ 0 \\ 1 \end{pmatrix}
       &&= \frac{1}{2}\begin{pmatrix} 1+1 \\ 0 \\ 1-1 \\ 0 \end{pmatrix}
        &= \begin{pmatrix} 1 \\ 0 \\ 0 \\ 0 \end{pmatrix}
       &&= \ket{00} \\
      %
      M\frac{\ket{01} + \ket{10}}{\sqrt{2}}
        &= M\frac{1}{\sqrt{2}}\begin{pmatrix} 0 \\ 1 \\ 1 \\ 0 \end{pmatrix}
       &&= \frac{1}{2}\begin{pmatrix} 0 \\ 1+1 \\ 0 \\ 1-1 \end{pmatrix}
        &= \begin{pmatrix} 0 \\ 1 \\ 0 \\ 0 \end{pmatrix}
       &&= \ket{01} \\
      %
      M\frac{\ket{00} - \ket{11}}{\sqrt{2}}
        &= M\frac{1}{\sqrt{2}}\begin{pmatrix} 1 \\ 0 \\ 0 \\ -1 \end{pmatrix}
       &&= \frac{1}{2}\begin{pmatrix} 1-1 \\ 0 \\ 1+1 \\ 0 \end{pmatrix}
        &= \begin{pmatrix} 0 \\ 0 \\ 1 \\ 0 \end{pmatrix}
       &&= \ket{10} \\
      %
      M\frac{\ket{01} - \ket{10}}{\sqrt{2}}
        &= M\frac{1}{\sqrt{2}}\begin{pmatrix} 0 \\ 1 \\ -1 \\ 0 \end{pmatrix}
       &&= \frac{1}{2}\begin{pmatrix} 0 \\ 1-1 \\ 0 \\ 1+1 \end{pmatrix}
        &= \begin{pmatrix} 0 \\ 0 \\ 0 \\ 1 \end{pmatrix}
       &&= \ket{11}
    \end{alignat*}
\end{enumerate}

In other words, teleportation boils down to the following: $T\ket{ψ00} = v'$,
where if we measure the first two bits of $v'$ to get $\hat{v}' \tensor
\ket{ψ'}$, we have $F_{\hat{v}'}\ket{ψ'} = \ket{ψ}$.

\end{document}

% ASZ: Notes that I want to use later:

% The set of qubits is ℚ; this is ambiguous, so we explain it.
, which we denote $\qubit$.\footnote{We do not have much
call to refer to rational numbers, so we use $\qubit$ to refer exclusively to
qubits, and never to rational numbers.}

% The real definition of a simple mixed state
we write this \[ \frac{1}{\sqrt{2}}\ket{0} + \frac{1}{\sqrt{2}}\ket{1} =
\frac{\ket{0} + \ket{1}}{\sqrt{2}}. \]

% ASZ: I use this to compile my LaTeX document; you can ignore it
(defun my-compile ()
  (interactive)
  (save-buffer)
  (shell-command (concat
    "pdflatex -halt-on-error quantum.tex && "
    "osascript ~/misc/Skim-open-revert.scpt quantum.pdf &")))
(local-set-key (kbd "C-c C-c") #'my-compile)
(local-set-key (kbd "C-c C-l") #'my-compile)
